Spatial population structure determines extinction risk in climate-induced range shifts

### 2019-06-28 Manuscript/AmNat_ReSubmission.tex
Revised the manuscript for resubmission

Previous models have predicted that dispersal evolution may be capable of rescuing populations that otherwise would be unable to keep pace with climate change~\citep{boeye2013more, henry2013eco}. While our results generally agree with this prediction, comparison with results from previous models suggests an important role for the biological and environmental context of dispersal evolution in range shifts. In a model assuming asexual reproduction, adaptation to an environmental gradient increased dispersal evolution during range shifts compared to scenarios with no gradient, contrary to the results presented here~\citep{hargreaves2015fitness}. In sexually reproducing populations, however, the evolution of increased dispersal can have an antagonistic interaction with local adaptation to an environmental gradient. Increased dispersal, and therefore increased gene flow, can reduce adaptation to the gradient via gene swamping~\citep{lenormand2002gene, kirkpatrick1997evolution}. When increased dispersal is necessary to track changing environmental conditions, our results suggest the fitness costs to the population can be severe, contributing to increased extinction risks in the presence of an environmental gradient. Thus, while dispersal evolution may be able to improve a population's ability to track changing conditions, it may also incur a severe cost depending on the environmental context of the population. For the purposes of this investigation, we assumed the environmental gradient shifted spatially with climate change, as would be expected if it corresponded to a gradient in temperature or precipitation~\citep{davis2001range}. However, in systems characterized by a gradient defined by other factors (e.g. geography or biotic interactions), the gradient might remain stable or even shift in an opposing direction to climate change. Future research should investigate the impact of an environmental gradient unrelated to climate on extinction risk during range shifts.

@@

\paragraph{Environmental gradient}
The local optimum for the niche trait ($z_{opt,x}$) varied in space according to
\begin{equation}
z_{opt,x}=\lambda(x-\beta_{t})
\end{equation}
with $\lambda$ determining the rate of change in the optimum across the range. Individual relative fitness ($w_{i,x}$) values were then calculated according to the following equation assuming stabilizing selection
\begin{equation}
w_{i,x}=e^{\frac{-(z_{i}-z_{opt,x})^{2}}{2\omega^{2}}}
\end{equation}
where $\omega$ defined the strength of stabilizing selection and $z_{i}$ was an individual's niche phenotype~\citep{lande1976natural}. Thus, an individual's realized fitness was higher the closer its niche phenotype ($z_{i}$) was to the environmental gradient value of the patch it occupied ($z_{opt,x}$). All loci contributed additively to an individual's niche value with no dominance or epistasis, meaning an individual's phenotype was simply the sum of the individual's allele values. As defined above, $z_{opt,x}$ also shifted with climate change (i.e. with $\beta_{t}$) as would be expected if it corresponded to a phenotypic optimum along a temperature or precipitation gradient within the range~\citep{davis2001range}.

\paragraph{Population dynamics}
Population growth within each patch was modeled with a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}. To account for fitness effects on population growth, expected population growth was scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$) so that maladaptation resulted in reduced population growth. The expected number of new offspring in patch $x$ at time $t+1$ was given by
\begin{equation}
\hat{N}_{t+1,x}=\bar{w_{x}}F_{t,x}\frac{R}{\psi}e^{\frac{-RN_{t,x}}{K_{x}}}
\end{equation}
where $F_{t,x}$ was the number of females in patch $x$ at time $t$, $R$ was the intrinsic growth rate for the population and remained constant in both time and space, $\psi$ was the expected sex ratio of the population, $N_{t,x}$ was the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ was the local carrying capacity based on the environmental conditions. To incorporate demographic stochasticity, the realized number of offspring for each patch was then drawn from a Poisson distribution.
\begin{equation}
N_{t+1,x}\sim Poisson(\hat{N}_{t+1,x})
\end{equation}

@@

% Population dynamics A3
\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
    \hline
    Parameter & Description & Value \\ \hline \hline
    $R$ & Intrinsic growth rate of the population & $2$ \\
    $\psi$ & Expected sex ratio (proportion of females) in the population & $0.5$ \\
    $\hat{d}$ & Maximum achievable dispersal phenotype & $1000$ \\
    $\rho$ & Determined the slope of the transition in dispersal phenotypes from $0$ to $\hat{d}$ & $0.5$ \\
    \hline
  \end{tabular}
\caption[LoF entry]{Values and descriptions for parameters related to population growth and dispersal.}
\label{table:PopPars}
\end{table}

@@

% Genetics A4
\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
    \hline
    Parameter & Description & Value \\ \hline \hline
    $\omega$ & Defined the strength of stabilizing selection on the niche trait & $3$ \\
    $U^{T}$ & Diploid mutation rate for each trait & $0.02$ \\
    $V_{m}^{T}$ & Mutational variance for each trait & $0.0004$ \\
    $L^{T}$ & Number of diploid loci defining each trait & $5$ loci \\
    $\mu_{1}^{f}$ & Initial mean allele value for the niche trait & $0$ \\
    $\mu_{1}^{d}$ & Initial mean allele value for the dispersal trait & $-1$ \\
    $\sigma_{1}^{f}$ & Initial standard deviation of allele values for the niche trait & $0.025$ \\
    $\sigma_{1}^{d}$ & Initial standard deviation of allele values for the dispersal trait & $1$ \\
    \hline
  \end{tabular}
\caption[LoF entry]{Values and descriptions for parameters defining the genetic components of the model.}
\label{table:GenPars}
\end{table}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/SchematicFigures/SimExample"}
\caption{A single example of a simulation with a steep environmental gradient and a moderately stark range edge. Information on the (a) abundance, (b) dispersal ability, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to $40$ generations after the start of climate change. Log transformed mean dispersal phenotypes (b) are shown for each patch. Average patch fitness (c) was calculated based on the mean niche trait of local individuals and the gradient value at each patch.}
\label{fig:SimExample}
\end{figure}

>>>>

Previous models have predicted that dispersal evolution may be capable of rescuing populations that otherwise would be unable to keep pace with climate change~\citep{boeye2013more, henry2013eco}. While our results generally agree with this prediction, comparison with results from previous models suggests an important role for the biological and environmental context of dispersal evolution in range shifts. In a model assuming asexual reproduction, adaptation to an environmental gradient increased dispersal evolution during range shifts compared to scenarios with no gradient~\citep{hargreaves2015fitness}, contrary to the results presented here. In sexually reproducing populations, however, the evolution of increased dispersal can have an antagonistic interaction with local adaptation to an environmental gradient. Increased dispersal, and therefore increased gene flow, can reduce adaptation to the gradient via gene swamping~\citep{lenormand2002gene, kirkpatrick1997evolution}. When increased dispersal is necessary to track changing environmental conditions, our results suggest the fitness costs to the population can be severe, contributing to increased extinction risks in the presence of an environmental gradient. Thus, while dispersal evolution may be able to improve a population's ability to track changing conditions, it may also incur a severe cost depending on the environmental context of the population. For the purposes of this investigation, we assumed the environmental gradient shifted spatially with climate change, as would be expected if it corresponded to a gradient in temperature or precipitation~\citep{davis2001range}. However, in systems characterized by a gradient defined by other factors (e.g. geography or biotic interactions), the gradient might remain stable or even shift in an opposing direction to climate change. Future research should investigate the impact of an environmental gradient unrelated to climate on extinction risk during range shifts.

@@

\paragraph{Environmental gradient}
The local optimum for the niche trait ($z_{opt,x}$) varied in space according to
\begin{equation}
z_{opt,x}=\lambda(x-\beta_{t})
\end{equation}
with $\lambda$ determining the rate of change in the optimum across the range. Individual relative fitness ($w_{i,x}$) values were then calculated according to the following equation assuming stabilizing selection
\begin{equation}
w_{i,x}=e^{\frac{-(z_{i}-z_{opt,x})^{2}}{2\omega^{2}}}
\end{equation}
where $\omega$ defined the strength of stabilizing selection and $z_{i}$ was an individual's niche phenotype~\citep{lande1976natural}. Thus, an individual's realized fitness was higher the closer its niche phenotype ($z_{i}$) was to the trait optimum at the patch it occupied ($z_{opt,x}$). All loci contributed additively to an individual's niche value with no dominance or epistasis, meaning an individual's phenotype was simply the sum of the individual's allele values. As defined above, $z_{opt,x}$ also shifted with climate change (i.e. with $\beta_{t}$) as would be expected if it corresponded to a phenotypic optimum along a temperature or precipitation gradient within the range~\citep{davis2001range}.

\paragraph{Population dynamics}
Population growth within each patch was modeled with a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}. To account for fitness effects on population growth, expected population growth was scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$) so that maladaptation resulted in reduced population growth. The expected number of new offspring in patch $x$ at time $t+1$ was given by
\begin{equation}
\hat{N}_{t+1,x}=\bar{w_{x}}F_{t,x}\frac{R}{\psi}e^{\frac{-RN_{t,x}}{K_{x}}}
\end{equation}
where $F_{t,x}$ was the number of females in patch $x$ at time $t$, $R$ was the intrinsic growth rate for the population and remained constant in both time and space, $\psi$ was the expected proportion of females in the population, $N_{t,x}$ was the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ was the local carrying capacity based on the environmental conditions. To incorporate demographic stochasticity, the realized number of offspring for each patch was then drawn from a Poisson distribution.
\begin{equation}
N_{t+1,x}\sim Poisson(\hat{N}_{t+1,x})
\end{equation}

@@

% Population dynamics A3
\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
    \hline
    Parameter & Description & Value \\ \hline \hline
    $R$ & Intrinsic growth rate of the population & $2$ \\
    $\psi$ & Expected proportion of females in the population & $0.5$ \\
    $\hat{d}$ & Maximum achievable dispersal phenotype & $1000$ \\
    $\rho$ & Determined the slope of the transition in dispersal phenotypes from $0$ to $\hat{d}$ & $0.5$ \\
    \hline
  \end{tabular}
\caption[LoF entry]{Values and descriptions for parameters related to population growth and dispersal.}
\label{table:PopPars}
\end{table}

@@

% Genetics A4
\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
    \hline
    Parameter & Description & Value \\ \hline \hline
    $\omega$ & Defined the strength of stabilizing selection on the niche trait & $3$ \\
    $U^{T}$ & Diploid mutation rate for each trait & $0.02$ \\
    $V_{m}^{T}$ & Mutational variance for each trait & $0.0004$ \\
    $L^{T}$ & Number of diploid loci defining each trait & $5$ loci \\
    $\mu_{1}^{n}$ & Initial mean allele value for the niche trait & $0$ \\
    $\mu_{1}^{d}$ & Initial mean allele value for the dispersal trait & $-1$ \\
    $\sigma_{1}^{n}$ & Initial standard deviation of allele values for the niche trait & $0.025$ \\
    $\sigma_{1}^{d}$ & Initial standard deviation of allele values for the dispersal trait & $1$ \\
    \hline
  \end{tabular}
\caption[LoF entry]{Values and descriptions for parameters defining the genetic components of the model.}
\label{table:GenPars}
\end{table}

@@

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/SchematicFigures/SimExample"}
\caption{A single example of a simulation with a steep environmental gradient and a moderately stark range edge. Information on the (a) abundance, (b) dispersal ability, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to $40$ generations after the start of climate change. Log transformed mean dispersal phenotypes (b) are shown for each patch. Average patch fitness (c) was calculated based on the mean niche trait of local individuals and the trait optimum at each patch.}
\label{fig:SimExample}
\end{figure}
